Retrofitting Effect Handlers onto OCaml

### 2020-10-31 retro-concurrency.tex
Updates
tabs to spaces
update static semantics

Our work is also timely. WebAssembly~\ref{} community group is actively
considering effect handlers as one of the mechanism for supporting concurrency,
asynchronous I/O and generators~\ref{WasmPropoal}. Project Loom~\cite{} is an
OpenJDK project which adds virtual threads and delimited continuations to Java.
We believe that our design choices will inform similar choices to be made in
the other languages.

@@

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b

type 'a computation = unit -> 'a
type ('a,'b) handler =
{ret: 'a -> 'b;
 eff_1: 'c1.'c1 eff -> ('c1,'b)continuation -> 'b;
 ...
 eff_n: 'cn.'cn eff -> ('cn,'b)continuation -> 'b}
val match_with: 'a computation -> ('a,'b) handler
                               -> 'b
\end{lstlisting}

We introduce the type |'a eff| of effect values which when performed using the
|perform| primitive returns a |'a| value. Constructors for the value of type
|'a eff| are declared using the effect declarations. For example, with the
declaration |effect E : string -> int|, the effect value |E "42"| has the type
|int eff|. We introduce the type |('a,'b) continuation| of delimited
continuations which expect a |'a| value for resumption and return |'b| value.
The continuations may be |continue|d with a suitably typed value or
|discontinue|d with an exception.

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. For the sake of exposition, we introduce a
|('a,'b) handler| type. This handler handles a |'a computation| that returns a
|'a| value, and itself returns a |'b| value. The handler has a single return
case of type |'a -> 'b|, and one or more effect cases. Each effect case,
handles a effect of type |'c eff|, whose continuation is of type
|('c,'b) continuation|, and returns a value of type |'b|. For example, in the
following code:
\begin{lstlisting}
effect E : string -> int
effect F : float -> int
let r = match perform (E "42") with
| v -> v == 0
| effect (E s) k ->
    not (continue k (int_of_string s))
| effect (F f) k ->
    not (continue k (int_of_float f))
\end{lstlisting}
\noindent |r| has type |bool|.

@@

\begin{figure}
	\begin{mathpar}
		\begin{array}{rc}

			\textsc{StepC} & \inferrule
			                    {(\term, \env, \fl, \ostack) \cstep \config}
													{\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} \\ \\
			\textsc{StepO} & \inferrule
													{(\term, \env, k, \cstack) \ostep \config}
													{\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
		\end{array}
	\end{mathpar}
	\caption{Top-level Reductions -- $\config \step \config$.}
	\label{sem:top}
\end{figure}

@@

\begin{figure*}
	\begin{smathpar}
		\begin{array}{rrcll}
			\textsc{AdminC}   & (\term,\env,\fl,\ostack) & \cstep
												& \configa{\term'}{\env'}{\cstacka{\fl'}{\ostack}}
												& \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
			\textsc{RetToO}   & (v,\env,[],\ostacka{k}{\cstack}) & \cstep
			                  & \configa{v}{\env}{\ostacka{k}{\cstack}} \\
			\textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep
												& \configa{e}{\envext{\env}{x}{v}}{\cstacka{\fl}{\ostack}} \\
			\textsc{Callback} & (v, \_, \oclos{x}{e}{\env}::\fl, \ostack) & \cstep
												& \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstacka{\fl}{\ostack}}}
												& \text{if } k = [[],(\{\caseval{x}{x}\},\emptyset)] \\
			\textsc{ExnFwdO}  & (v, \env, \ffun{\exnval{l}}::\_, \ostacka{(\fl,\hc) \kcons k}{\cstack}) & \cstep
												& \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl,\hc) \kcons k}{\cstack}}
		\end{array}
	\end{smathpar}
	\caption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
	\label{sem:cstep}
\end{figure*}

\begin{figure*}
	\begin{smathpar}
		\begin{array}{rrcl}
			\textsc{Var}     & (x,\env, \fl) & \rightsquigarrow
			                 & (\env(x), \env, \fl) \\
			\textsc{Arith1}  & (e_1 \odot e_2, \env, \fl) & \rightsquigarrow
			                 & (e_1,\env, \faritha{\odot}{e_2}{\env}::\fl) \\
			\textsc{Arith2}  & (n_1, \_, \faritha{\odot}{e_2}{\env}::\fl) & \rightsquigarrow
			                 & (e_2, \env, \farithb{\odot}{n_1}::\fl) \\
			\textsc{Arith3}  & (n_2, \env, \farithb{\odot}{n_1}::\fl) & \rightsquigarrow
			                 & (\llbracket n_1 \odot n_2 \rrbracket, \env, \fl) \\
			\textsc{App1}    & (e_1 ~e_2, \env, \fl) & \rightsquigarrow
			                 & (e_1, \env, \farg{e_2}{\env}::\fl) \\
			\textsc{App2}    & (\lam{x}{e}, \env, \fl) & \rightsquigarrow
			                 & (\clos{x}{e}{\env}, \env, \fl) \\
			\textsc{App3}    & (\clos{x}{e_1}{\env_1}, \_, \farg{e_2}{\env_2}::\fl) & \rightsquigarrow
			                 & (e_2, \env_2, \ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
			\textsc{Resume1} & (k,\_,\farg{e_1}{\env_1}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
			                 & (e_1, \env_1, \ffun{k}::\farg{e_2}{\env_2}::\fl) \\
			\textsc{Resume2} & (\clos{x}{e_1}{\env_1}, \_, \ffun{k}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
			                 & (e_2, \env_2, \ffun{k}::\ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
			\textsc{Perform} & (\perform{l}{e}, \env, \fl) & \rightsquigarrow
			                 & (e, \env, \ffun{\effval{l}{[[],(\{\caseval{x}{x}\},\emptyset)]}}::\fl) \\
			\textsc{Raise}   & (\throw{l}{e}, \env, \fl) & \rightsquigarrow
											 & (e, \env, \ffun{\exnval{l}}::\fl)
		\end{array}
	\end{smathpar}
	\caption{Administrative Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
	\label{sem:step}
\end{figure*}

>>>>

Our work is also timely. WebAssembly~\ref{} community group is actively
considering effect handlers as one of the mechanism for supporting concurrency,
asynchronous I/O and generators~\ref{WasmPropoal}. Project Loom~\cite{} is an
OpenJDK project which adds virtual threads and delimited continuations to Java.
Swift concurrency roadmap~\cite{} includes direct-style asynchronous
programming and structured concurrency. We believe that our design choices will
inform similar choices to be made in the other languages.

@@

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff = ..
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b

(* Internal API *)
type 'a comp = unit -> 'a
type ('a,'b) handler =
{retc: 'a -> 'b;
 effc: 'c.'c eff -> ('c,'b)continuation -> 'b; }
val match_with: 'a comp -> ('a,'b) handler -> 'b
\end{lstlisting}

We introduce an extensible variant type~\cite{} |'a eff| of effect values,
which when performed using the |perform| primitive returns a |'a| value.
Constructors for the value of type |'a eff| are declared using the effect
declarations. For example, the declaration |effect E : string -> int| is
syntactic sugar for adding a new constructor to the variant type
|type _ eff += E : string -> int eff|. We introduce the type
|('a,'b) continuation| of delimited continuations which expect a |'a| value for
resumption and return |'b| value. The continuations may be |continue|d with a
suitably typed value or |discontinue|d with an exception.

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
    | (E s) -> (fun k1 -> e1)
    | (F f) -> (fun k2 -> e2)
    | e -> (fun k -> match perform e with
            | v -> continue k v
            | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return case of type |'a -> 'b|. The effect case |effc|
handles effects of type |'c eff| with |('c,'b) continuation| and returns a
value of type |'b|. The last pattern in |effc| forwards any unhandled effect to
the outer handler and returns the value and exceptions back.

@@

\if{0}
\begin{figure}
  \begin{mathpar}
    \begin{array}{rc}
      \textsc{StepC} & \inferrule
                          {(\term, \env, \fl, \ostack) \cstep \config}
                          {\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} \\ \\
      \textsc{StepO} & \inferrule
                          {(\term, \env, k, \cstack) \ostep \config}
                          {\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
    \end{array}
  \end{mathpar}
  \caption{Top-level Reductions -- $\config \step \config$.}
  \label{sem:top}
\end{figure}
\fi

\begin{figure}
  \begin{mathpar}
    \begin{array}{cc}
      \textsc{StepC} ~
      \inferrule{(\term, \env, \fl, \ostack) \cstep \config}
                {\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} &
      \textsc{StepO} ~
      \inferrule{(\term, \env, k, \cstack) \ostep \config}
                {\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
    \end{array}
  \end{mathpar}
  \caption{Top-level Reductions -- $\config \step \config$.}
  \label{sem:top}
\end{figure}

@@

\begin{figure*}
  \begin{smathpar}
    \begin{array}{rrcl}
      \textsc{Var}     & (x,\env, \fl) & \rightsquigarrow
                       & (\env(x), \env, \fl) \\
      \textsc{Arith1}  & (e_1 \odot e_2, \env, \fl) & \rightsquigarrow
                       & (e_1,\env, \faritha{\odot}{e_2}{\env}::\fl) \\
      \textsc{Arith2}  & (n_1, \_, \faritha{\odot}{e_2}{\env}::\fl) & \rightsquigarrow
                       & (e_2, \env, \farithb{\odot}{n_1}::\fl) \\
      \textsc{Arith3}  & (n_2, \env, \farithb{\odot}{n_1}::\fl) & \rightsquigarrow
                       & (\llbracket n_1 \odot n_2 \rrbracket, \env, \fl) \\
      \textsc{App1}    & (e_1 ~e_2, \env, \fl) & \rightsquigarrow
                       & (e_1, \env, \farg{e_2}{\env}::\fl) \\
      \textsc{App2}    & (\lam{x}{e}, \env, \fl) & \rightsquigarrow
                       & (\clos{x}{e}{\env}, \env, \fl) \\
      \textsc{App3}    & (\clos{x}{e_1}{\env_1}, \_, \farg{e_2}{\env_2}::\fl) & \rightsquigarrow
                       & (e_2, \env_2, \ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
      \textsc{Resume1} & (k,\_,\farg{e_1}{\env_1}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
                       & (e_1, \env_1, \ffun{k}::\farg{e_2}{\env_2}::\fl) \\
      \textsc{Resume2} & (\clos{x}{e_1}{\env_1}, \_, \ffun{k}::\farg{e_2}{\env_2}::\fl) & \rightsquigarrow
                       & (e_2, \env_2, \ffun{k}::\ffun{\clos{x}{e_1}{\env_1}}::\fl) \\
      \textsc{Perform} & (\perform{l}{e}, \env, \fl) & \rightsquigarrow
                       & (e, \env, \ffun{\effval{l}{[[],(\{\caseval{x}{x}\},\emptyset)]}}::\fl) \\
      \textsc{Raise}   & (\throw{l}{e}, \env, \fl) & \rightsquigarrow
                       & (e, \env, \ffun{\exnval{l}}::\fl)
    \end{array}
  \end{smathpar}
  \caption{Administrative Reductions -- $(\term, \env, \fl) \rightsquigarrow (\tau, \env, \fl)$.}
  \label{sem:step}
\end{figure*}
\begin{figure*}
  \begin{smathpar}
    \begin{array}{rrcll}
      \textsc{AdminC}   & (\term,\env,\fl,\ostack) & \cstep
                        & \configa{\term'}{\env'}{\cstacka{\fl'}{\ostack}}
                        & \text{if } (\term,\env,\fl) \rightsquigarrow (\term',\env',\fl') \\
      \textsc{RetToO}   & (v,\env,[],\ostacka{k}{\cstack}) & \cstep
                        & \configa{v}{\env}{\ostacka{k}{\cstack}} \\
      \textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep
                        & \configa{e}{\envext{\env}{x}{v}}{\cstacka{\fl}{\ostack}} \\
      \textsc{Callback} & (v, \_, \oclos{x}{e}{\env}::\fl, \ostack) & \cstep
                        & \configa{e}{\envext{\env}{x}{v}}{\ostacka{k}{\cstacka{\fl}{\ostack}}}
                        & \text{if } k = [[],(\{\caseval{x}{x}\},\emptyset)] \\
      \textsc{ExnFwdO}  & (v, \env, \ffun{\exnval{l}}::\_, \ostacka{(\fl,\hc) \kcons k}{\cstack}) & \cstep
                        & \configa{v}{\env}{\ostacka{(\ffun{\exnval{l}}::\fl,\hc) \kcons k}{\cstack}}
    \end{array}
  \end{smathpar}
  \caption{C Reductions -- $(\term, \env, \fl, \ostack) \cstep \config$.}
  \label{sem:cstep}
\end{figure*}
